Concurrent Programming with CXL

### 2020-06-14 book/paper.tex
Latest

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes concurrently read and update shared variables and
synchronize with one another makes programs more complicated than
programs where only one thing happens at a time.  Why is this?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential program is mostly \emph{deterministic}.
If you run it twice with the same input, the same things will happen.
Bugs are typically easily reproducible.
\item Each statement and each function can be thought of as happening
\emph{atomically} because there is no other activity interfering with
their execution.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

@@

So why is this useful?  Consider, for example, a sorting algorithm.
Now suppose we create a test program, a model, that tries sorting
\emph{all} lists of up to five numbers chosen from the set \{ 1,
2, 3, 4, 5 \}.  Model checking proves that for those particular
scenarios the sorting algorithm works: the output is a sorted
permutation of the input.  In some sense it is an excellent test:
it will have considered all corner cases, including lists where all
numbers are the same, lists that are already sorted or reversely
sorted, etc.  If there is a bug in the sorting algorithm, most
likely it would be triggered and the model checker would produce a
scenario that would make it easy to find the source of the bug.
However, if the model checker does not find any bugs, we do not
know for sure that the algorithm works for lists with more than
five numbers or for lists that have values other than the numbers
1 through 5.  Still, we would expect that the likelihood that there
are bugs remaining in the sorting algorithm is small.

But model checking can also help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}: predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
those invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof, even an
informal one.
Throughout this book we will include examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

@@

The syntax and semantics of CXL is similar to that of Python.
Python is familiar to many programmers and is easy to learn and
use.  We will assume that the reader is familiar with the basics
of Python programming.  We also will assume that the reader
understand some basics of machine architecture and how programs
are executed.  For example, we assume that the reader is familiar
with the concepts of CPU, memory, register, and stack.

@@

Like Python, CXL is an imperative,
dynamically typed, and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Every statement in CXL must be terminated by a semicolon
(even \texttt{while} statements and \texttt{def} statements);
in Python semicolons are optional and rarely used;
\item Correct indentation in CXL is encouraged but optional;
\item Python is object-oriented; CXL is not;
\item CXL does not (currently) support floating point;
\item CXL does not support I/O of any kind;
\item CXL only supports basic operator precedence or associativity.
Use parentheses liberally to remove ambiguity.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

Figure~\ref{fig:triangle} gives a simple example of a CXL program.
It is sequential and has a method \texttt{square} that takes
an integer number as argument.  Each method has a variable called
\texttt{result} that eventually contains the result of the
method (there is no \texttt{return} statement in CXL).  The method
also has a variable called \texttt{n} containing the value of the
argument.  The \texttt{x..y} notation generates a set containing the numbers
from~\texttt{x} to~\texttt{y} (inclusive).  The last two lines in the program are
the most interesting.
They assign to \texttt{x} some unspecified value in the range \texttt{0..N}
and then verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.

@@

\begin{quote}
What will happen?
\begin{itemize}
\item Before you run the program, what do you think will happen?  Is the
program correct in that \texttt{count} will always end up being 2?
\item What are the possible values of \texttt{count} at the time of the
\texttt{assert} statement?
(You may assume that \texttt{load} and \texttt{store} instructions of the
underlying architecture are atomic---in fact they are.)
\end{itemize}
\end{quote}

@@

The following is a possible interleaving of incrementers~0 and~1:
\begin{enumerate}
\item incrementer 0 loads the value of \texttt{count}, which is 0;
\item incrementer 1 loads the value of \texttt{count}, which is still 0;
\item incrementer 1 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item incrementer 0 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item incrementer 0 sets \texttt{upDone} to \texttt{True};
\item incrementer 1 sets \texttt{upDone} to \texttt{True}.
\end{enumerate}

@@

CXL makes extensive use of dictionaries.  Their syntax and properties
are a little different from Python.
A dictionary maps CXL values, known as \emph{keys}, to CXL values.
Unlike Python, any CXL value can be a key, including another
dictionary.
CXL dictionaries are written as
$\mathtt{dict}\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
If \texttt{d} is a dictionary, and \texttt{k} is a key, then the
following expression retrieves the CXL value that $k$ maps to:
\begin{code}
\begin{verbatim}
d k
\end{verbatim}
\end{code}
This is unfamiliar to Python programers, but in CXL square brackets can be used
in the same way as parentheses, so you can express it in the more familiar form:
\begin{code}
\begin{verbatim}
d[k]
\end{verbatim}
\end{code}
However, if \texttt{d = dict\{ .count: 3 \}}, then you can write
can write \texttt{d.count} (which has value 3) instead of having to write
\texttt{d[.count]} (although both will work).
The meaning of \texttt{d a b c ...} is \texttt{((((d a) b) c) ...)}.

@@

Before we delve into how to solve synchronization problems, it is important
to know a bit about the underlying machine.  A CXL program is translated
into a list of machine instructions that the CXL machine executes.
The CXL machine is not an ordinary virtual machine, but its architecture
is nonetheless representative of conventional machines such as hardware
with memory and CPUs and virtual machines such as the Java Virtual Machine.

@@

The state of a process is called a \emph{context}: it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag}) of
contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty dictionary, \texttt{()}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
CXL values are unbounded trees.
Both the memory and the register of a process always contain
a dictionary that maps atoms to CXL values.  We call this a \emph{directory}.
A directory represents the state of a collection of variables named by the atoms.
%
Because directories are CXL values themselves,
directories can be organized into a tree.
Each node in the tree is then identified
by a sequence of atoms, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address} of a CXL value, and it is relative to the
``root'' of the CXL value.

Compiling the code in Figure~\ref{fig:inc} results in the CXL machine code
listed in Figure~\ref{fig:inccode}.
You can obtain this code by invoking \texttt{cxl} with the \texttt{-a} flag
like so:
\begin{code}
\begin{verbatim}
cxl -a Up.cxl
\end{verbatim}
\end{code}
The CXL machine is predominantly a \emph{stack machine}.
All instructions are atomically executed.
Most instructions pop values from the stack or push values onto the stack.
Initially there is one process (with name tag \texttt{\_\_init\_\_/()}
that starts executing at instruction 0 and keeps executing until the last
execution in the program to initialize the state.
In this case, the first instruction is a \texttt{JUMP} instruction that sets the
program counter to 39.
At program counter 1 is the code for the \texttt{incrementer} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

The \texttt{Frame} instruction has three arguments:
\begin{enumerate}
\item The name of the method;
\item A list containing the names of the arguments of the method;
\item The program counter of the method's \texttt{Return} instruction.
\end{enumerate}

The code generated from \texttt{count := count + 1} in line 2 of
\texttt{Up.cxl} is as follows:

@@

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A counter is a very simple example of
a data structure.  A more involved one would be access to a binary tree.
Adding a node to a binary tree, or re-balancing a tree, often requires
multiple operations.  Maintaining ``consistency'' is certainly much easier
(although not necessarily impossible) if during this time no other
process also tries to access the binary tree.

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CXL code.
Here \texttt{@ncs} and \texttt{@cs} are \emph{labels}, each identifying
a location in the CXL machine code.  The first thing we need to
ensure is that there cannot be two processes at the critical section.
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.  CXL in fact
supports this.  It has an operator \texttt{atLabel $L$}, where $L$
is the atom of the label (in this case, \texttt{.cs}).  The
operator returns a bag of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.
Method \texttt{atLabel} only exists for specification purposes---do not
use it in normal code.

The assertion also makes use of the \texttt{nametag()} operator that
returns the name tag of the current process.
If you run the code through CXL, the assertion should fail because
there is no code yet for entering and exiting the critical section.

@@

@@

In order to detect violations of progress, and other liveness problems in
algorithms in general, CXL requires that every execution must be
able to reach a state in which all processes have terminated.
Clearly, even if mutual exclusion holds in Figure~\ref{fig:csbarebones},
the spawned processes never terminate.  In order to resolve this, we
will model processes in critical sections using the framework in
Figure~\ref{fig:cs}: a process can \emph{choose} to enter a
critical section more than once, but it can also choose to terminate, even
without entering the critical section ever.
Note also that we have dropped the \texttt{@ncs} label: it would serve no
specific purpose.
Figure~\ref{fig:mutex} show a high-level state diagram of the code in
Figure~\ref{fig:cs}.

@@

In order to have at most one process in a critical section at a time, you
probably are thinking that one might use a \emph{lock}.
That is a good thought, but how does one implement one?
Figure~\ref{fig:uplock} presents a mutual exclusion attempt based on a
na\"{\i}ve implementation of a lock.
The idea is that the lock is like a baton that at most one process can own
(or hold) at a time.
Initially the lock is not owned, indicated by lock being \texttt{False}.
To enter the critical section, a process waits until \texttt{lock} is \texttt{False}
and then sets it to \texttt{True} to indicate that the lock has been taken.
The process then executes the critical section.  Finally the process
releases the lock by setting it back to \texttt{False}.

Unfortunately, if we run the program, we find that the assertion still
fails.  Diagnosing the problem, we see that the \texttt{lock} variable
suffers from the same problem as the \texttt{count} variable
in Figure~\ref{fig:inc}: operations
on it consist of several instructions.  It is thus possible
for both processes to believe the lock is available and to obtain the lock
at the same time.

@@

To see why, first note the invariant that if process $i$ is in the
critical section, then flags[i] == True.
Without loss of generality,
suppose that process~0 sets \texttt{flags[0]} at time $t_0$.
Process 0 can only reach the critical section if at some time $t_1$,
$t_1 > t_0$, it finds that \texttt{flags[1] == False}.
Because of the invariant, \texttt{flags[1] == False} implies that
process~1 is not at the critical section at time $t_1$.
Let $t_2$ be the time at which process~0 sets \texttt{flags[0]}
to \texttt{False}.  Process~0 is in the critical section sometime
between $t_1$ and $t_2$.
It is easy to see that process~1 cannot enter the critical section
between $t_1$ and $t_2$, because \texttt{flags[1] == False} at
time $t_1$.  To reach the critical section between $t_1$ and $t_2$,
it would first have to set \texttt{flags[1]} to \texttt{True} and
then wait until \texttt{flags[0] == False}.  But that does not happen
until time $t_2$.

However, the solution has a different problem called \emph{livelock}:
if both try to enter the critical section at the same time, they may
end up waiting for one another indefinitely.  Thus the solution
violates \emph{progress}.

The final na\"{\i}ve solution is based on a variable called \texttt{turn}
that alternates between 0 and 1.  When \texttt{turn = $i$}, process~$i$ can
enter the critical section, while process $1-i$ has to wait.  When done,
process~$i$ sets \texttt{turn} to $1-i$ to give the other process an
opportunity to enter.
An invariant of this solution is that while process~$i$ is in the critical
section, \texttt{turn} == $i$.
Since \texttt{turn} cannot be 0 and 1 at
the same time, mutual exclusion is satisfied.
The solution also has the nice property that
processes 0 and 1 alternate entering the critical section.
The problem with the solution is
that it also violates the \emph{progress} property:
if process~$i$ terminates instead of entering the critical section when it
is process~$i$'s turn, process $1-i$ ends up waiting indefinitely for its
turn.

@@

In 1981, Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
Figures~\ref{fig:upflags} and~\ref{fig:upturn}, and is presented
in Figure~\ref{fig:peterson}.
A process first indicates its interest in entering the critical section
by setting its flag.
It then politely gives way to the other process should it also want to
enter the critical section---if both do so at the same time one will
win because writes to memory in CXL are atomic.
The process continues to be polite, waiting in a \texttt{while} loop
until either the other process is nowhere near the critical section
or has given way.
Running the algorithm with CXL shows that it satisfies both mutual
exclusion and progress.

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the CXL
machine works.
In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
at any point of time the CXL machine is in a specific state.
Everytime a process executes a CXL machine instruction, the
state changes (if only because the program counter of the process is
updated).  We call that a \emph{step}.  Steps in CXL are atomic.

The CXL machine starts in an initial state in which there is only
one process and its program counter is~0.  A \emph{trace} is a
sequence of steps starting from the initial state.  When making a
step, there are two sources of non-determinism in CXL.  One is when
in some state more than one process can make a step.  The other is
when a process executes a \texttt{choose} operation and there is
more than one choice.

A good way to prove mutual exclusion is through induction on
the number of steps.
In such a proof, we need a so-called \emph{inductive invariant} $I$ that
implies mutual exclusion.
An inductive invariant $I$ is a predicate over state that satisfies the following:
\begin{itemize}
\item $I$ hold in the initial state.
\item For any state in which $I$ holds and any process in the
state takes a step, then $I$ also holds in the resulting state.
\end{itemize}

@@

Let $C(i) = \mathtt{flags}[1 - i] \land
\mathtt{turn} = 1 - i$, that is, the condition on the \texttt{while} loop
for process~$i$.
Let predicate $I_p(i)$ be the following.
if process $i$ is label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $C(i)$ does not hold or process $1-i$ is executing after setting
$\mathtt{flags}[i]$ but still before setting \texttt{turn} to $1-i$.
More formally, $I_p(i) = \mathtt{process}(i)@cs) \Rightarrow (\lnot C(i) \lor \mathtt{process}(1-i)@gate)$.
For Peterson's Algorithm, an inductive invariant that works well is
$I_p(0) \land I_p(1)$.

>>>>

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes concurrently read and update shared variables and
synchronize with one another makes programs more complicated than
programs where only one thing happens at a time.  Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential program is mostly \emph{deterministic}.
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
The output of running concurrent programs depends on how the various
processes are interleaved.  Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically} because there is no other activity
interfering with their execution.  Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other instructions
may update memory locations as a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

@@

So if model checking does not prove a program correct, why is it
useful?  To answer that question consider, for example, a sorting
algorithm.
Now suppose we create a test program, a model, that tries sorting
\emph{all} lists of up to five numbers chosen from the set \{ 1,
2, 3, 4, 5 \}.  Model checking proves that for those particular
scenarios the sorting algorithm works: the output is a sorted
permutation of the input.  In some sense it is an excellent test:
it will have considered all corner cases, including lists where all
numbers are the same, lists that are already sorted or reversely
sorted, etc.  If there is a bug in the sorting algorithm, most
likely it would be triggered and the model checker would produce a
scenario that would make it easy to find the source of the bug.
However, if the model checker does not find any bugs, we do not
know for sure that the algorithm works for lists with more than
five numbers or for lists that have values other than the numbers
1 through 5.  Still, we would expect that the likelihood that there
are bugs remaining in the sorting algorithm is small.
Note, however, that it would be easy to write a program
that sorts all lists of up to five numbers correctly but fails to
do so for a list of 6 numbers.  (Hint: simply use an \texttt{if}
statement.)

While model checking does not in general prove an algorithm correct,
it can help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}: predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
proposed invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof, even an
informal one.
Throughout this book we will include examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

@@

The syntax and semantics of CXL is similar to that of Python.
Python is familiar to many programmers and is easy to learn and
use.  We will assume that the reader is familiar with the basics
of Python programming.  We also will assume that the reader
understand some basics of machine architecture and how programs
are executed.  For example, we assume that the reader is familiar
with the concepts of CPU, memory, register, stack, and machine
instructions.

@@

Like Python, CXL is an imperative,
dynamically typed, and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Syntax: Every statement in CXL must be terminated by a semicolon
(even \texttt{while} statements and \texttt{def} statements).
In Python semicolons are optional and rarely used.
There is no syntactic significance to indentation in CXL.
Correct indentation in CXL is encouraged but not enforced.
CXL only supports basic operator precedence or associativity.
Use parentheses liberally to remove ambiguity.
\item CXL does not (currently) support floating point.  CXL does not support I/O;
\item Python is object-oriented; CXL is not;
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

Figure~\ref{fig:triangle} gives a simple example of a CXL program.
It is sequential and has a method \texttt{square} that takes
an integer number as argument.  Each method has a variable called
\texttt{result} that eventually contains the result of the
method (there is no \texttt{return} statement in CXL).  The method
also has a variable called \texttt{n} containing the value of the
argument.  The \texttt{x..y} notation generates a set containing the numbers
from~\texttt{x} to~\texttt{y} (inclusive).  The last two lines in the program are
the most interesting.
The first assigns to \texttt{x} some unspecified value in the range \texttt{0..N}
and the second verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.

@@

\begin{quote}
What will happen?
\begin{itemize}
\item Before you run the program, what do you think will happen?  Is the
program correct in that \texttt{count} will always end up being 2?
\item What are the possible values of \texttt{count} at the time of the
\texttt{assert} statement?
(You may assume that \texttt{load} and \texttt{store} instructions of the
underlying machine architecture are atomic---in fact they are.)
\end{itemize}
\end{quote}

@@

The following is a possible interleaving of incrementers~0 and~1:
\begin{enumerate}
\item incrementer 0 loads the value of \texttt{count}, which is 0;
\item incrementer 1 loads the value of \texttt{count}, which is still 0;
\item incrementer 1 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item incrementer 0 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item incrementer 0 sets \texttt{done[0]} to \texttt{True};
\item incrementer 1 sets \texttt{done[1]} to \texttt{True}.
\end{enumerate}

@@

CXL makes extensive use of dictionaries.
A dictionary maps values, known as \emph{keys}, to values.
The CXL dictionary syntax and properties are a little different from Python.
Unlike Python, any CXL value can be a key, including another
dictionary.
CXL dictionaries are written as
$\mathtt{dict}\{ k_0: v_0, ~ k_1: v_1, ~ ... \}$.
If \texttt{d} is a dictionary, and \texttt{k} is a key, then the
following expression retrieves the CXL value that $k$ maps to:
\begin{code}
\begin{verbatim}
d k
\end{verbatim}
\end{code}
This is unfamiliar to Python programers, but in CXL square brackets can be used
in the same way as parentheses, so you can express the same thing in the form
that is familiar to Python programmers:
\begin{code}
\begin{verbatim}
d[k]
\end{verbatim}
\end{code}
However, if \texttt{d = dict\{ .count: 3 \}}, then you can write
can write \texttt{d.count} (which has value 3) instead of having to write
\texttt{d[.count]} (although both will work).
Thus using atoms, a dictionary can be made to look much like a Python object.
The meaning of \texttt{d a b c ...} is \texttt{((((d a) b) c) ...)}.

@@

Before we delve into how to solve synchronization problems, it is important
to know a bit about the underlying machine.  A CXL program is translated
into a list of machine instructions that the CXL machine executes.
The CXL machine is not an ordinary virtual machine, but its architecture
is nonetheless representative of conventional computers and
virtual machines such as the Java Virtual Machine.

@@

The state of a process is called a \emph{context}: it contains the value of
its name tag, program counter, stack, and register.
The state of the CXL machine
consists of the value of its memory and the multiset (or \emph{bag}) of
contexts.  It is a multiset of contexts because two different processes can
be in the same state.
The initial state of the CXL memory is the empty dictionary, \texttt{dict\{\}}.
The context bag has an initial context in it with name tag
\texttt{\_\_main\_\_/()}, pc 0, register \texttt{()}, and an empty stack.
Each machine instruction updates the state in some way.
Figure~\ref{fig:incstate} shows an example of a reachable state for 
the program in Figure~\ref{fig:inc}.

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
CXL values are unbounded trees.
Both the memory and the register of a process always contain
a dictionary that maps atoms to CXL values.  We call this a \emph{directory}.
A directory represents the state of a collection of variables named by the atoms.
%
Because directories are CXL values themselves,
directories can be organized into a tree.
Each node in a directory tree is then identified
by a sequence of atoms, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address} of a CXL value, and it is relative to the
``root'' of the directory tree.

Compiling the code in Figure~\ref{fig:inc} results in the CXL machine code
listed in Figure~\ref{fig:inccode}.
You can obtain this code by invoking \texttt{cxl} with the \texttt{-a} flag
like so:
\begin{code}
\begin{verbatim}
cxl -a Up.cxl
\end{verbatim}
\end{code}
The CXL machine is predominantly a \emph{stack machine}.
All instructions are atomically executed.
Most instructions pop values from the stack or push values onto the stack.
At first there is one process (with name tag \texttt{\_\_init\_\_/()}
that initializes the state.
It starts executing at instruction 0 and keeps executing until the last
execution in the program.
In this case, the first instruction is a \texttt{JUMP} instruction that sets the
program counter to 39.
At program counter 1 is the code for the \texttt{incrementer} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

The \texttt{Frame} instruction has three arguments: the name of the method,
a list containing the names of the arguments of the method, and
the program counter of the method's \texttt{Return} instruction.
The code generated from \texttt{count := count + 1} in line 2 of
\texttt{Up.cxl} is as follows:

@@

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A counter is a very simple example of
a data structure, but as we have seen it too requires multiple instructions.
A more involved one would be access to a binary tree.
Adding a node to a binary tree, or re-balancing a tree, often requires
multiple operations.  Maintaining ``consistency'' is certainly much easier
(although not necessarily impossible) if during this time no other
process also tries to access the binary tree.

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CXL code.
Here \texttt{@cs} is a \emph{label}, identifying
a location in the CXL machine code.  The first thing we need to
ensure is that there cannot be two processes at the critical section.
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

CXL in fact supports this.
It has an operator \texttt{atLabel $L$}, where $L$
is the atom containing the name of the label (in this case, \texttt{.cs}).
The operator returns a bag (multiset) of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.
Method \texttt{atLabel} only exists for specification purposes---do not
use it in normal code.
The assertion also makes use of the \texttt{nametag()} operator that
returns the name tag of the current process.
If you run the code through CXL, the assertion should fail because
there is no code yet for entering and exiting the critical section.

@@

In order to detect violations of progress, and other liveness problems in
algorithms in general, CXL requires that every execution must be
able to reach a state in which all processes have terminated.
Clearly, even if mutual exclusion holds in Figure~\ref{fig:csbarebones},
the spawned processes never terminate.  In order to resolve this, we
will model processes in critical sections using the framework in
Figure~\ref{fig:cs}: a process can \emph{choose} to enter a
critical section more than once, but it can also choose to terminate, even
without entering the critical section ever.

@@

Figure~\ref{fig:mutex} show a high-level state diagram of the code in
Figure~\ref{fig:cs}.
The circles represent states, while the arrows represent possible state
transitions.  The label in each circle summarizes the state;  the label
on an arrow describes the transition.  In this case,
the label contains two numbers: the total number of processes and the number
of processes in the critical section.  A process that is in the critical
section cannot terminate.

@@

You may already have heard of the concept of a \emph{lock} and have realized that
it could be used to implement a critical section.
The idea is that the lock is like a baton that at most one process can own
(or hold) at a time.
A process that wants to enter the critical section at a time must obtain the
lock first and release it upon exiting the critical section.
Using a lock is a good thought, but how does one implement one?
Figure~\ref{fig:uplock} presents a mutual exclusion attempt based on a
na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
Initially the lock is not owned, indicated by lock being \texttt{False}.
To enter the critical section, a process waits until \texttt{lock} is \texttt{False}
and then sets it to \texttt{True} to indicate that the lock has been taken.
The process then executes the critical section.  Finally the process
releases the lock by setting it back to \texttt{False}.

Unfortunately, if we run the program through CXL, we find that the assertion still
fails.  Diagnosing the problem, we see that the \texttt{lock} variable
suffers from the same problem as the \texttt{count} variable
in Figure~\ref{fig:inc}: operations
on it consist of several instructions.  It is thus possible
for both processes to believe the lock is available and to obtain the lock
at the same time.

@@

To see why, first note the invariant that if process $i$ is in the
critical section, then \texttt{flags[$i$] == True}.
Without loss of generality,
suppose that process~0 sets \texttt{flags[0]} at time $t_0$.
Process 0 can only reach the critical section if at some time $t_1$,
$t_1 > t_0$, it finds that \texttt{flags[1] == False}.
Because of the invariant, \texttt{flags[1] == False} implies that
process~1 is not at the critical section at time $t_1$.
Let $t_2$ be the time at which process~0 sets \texttt{flags[0]}
to \texttt{False}.  Process~0 is in the critical section sometime
between $t_1$ and $t_2$.
It is easy to see that process~1 cannot enter the critical section
between $t_1$ and $t_2$, because \texttt{flags[1] == False} at
time $t_1$.  To reach the critical section between $t_1$ and $t_2$,
it would first have to set \texttt{flags[1]} to \texttt{True} and
then wait until \texttt{flags[0] == False}.  But that does not happen
until time $t_2$.

Run the program through CXL.  Unfortunately, it turns out the solution has a
problem: if both try to enter the critical section at the same time, they may
end up waiting for one another indefinitely.  Thus the solution
violates \emph{progress}.

The final na\"{\i}ve solution is based on a variable called \texttt{turn}
that alternates between 0 and 1.  When \texttt{turn = $i$}, process~$i$ can
enter the critical section, while process $1-i$ has to wait.  When done,
process~$i$ sets \texttt{turn} to $1-i$ to give the other process an
opportunity to enter.
An invariant of this solution is that while process~$i$ is in the critical
section, \texttt{turn} == $i$.
Since \texttt{turn} cannot be 0 and 1 at
the same time, mutual exclusion is satisfied.
The solution also has the nice property that
processes 0 and 1 alternate entering the critical section.

Run the program through CXL.  It turns out that this solution also violates
\emph{progress}, albeit for a different reason:
if process~$i$ terminates instead of entering the critical section when it
is process~$i$'s turn, process $1-i$ ends up waiting indefinitely for its
turn.

\begin{quote}
Try your own solutions!
\begin{itemize}
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with CXL and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure it out---as
we will find out, it is possible but not easy.
\end{itemize}
\end{quote}

@@

In 1981, Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
Figures~\ref{fig:upflags} and~\ref{fig:upturn}, and is presented
in Figure~\ref{fig:peterson}.
A process first indicates its interest in entering the critical section
by setting its flag.
It then politely gives way to the other process should it also want to
enter the critical section---if both do so at the same time one will
win because writes to memory in CXL are atomic.
The process continues to be polite, waiting in a \texttt{while} loop
until either the other process is nowhere near the critical section
(\texttt{flag[1 - self] == False}) or or has given way
(\texttt{turn == self}).
Running the algorithm with CXL shows that it satisfies both mutual
exclusion and progress.

\begin{figure}
\begin{center}
\includegraphics[width=4in]{figures/states.jpeg}
\end{center}
\caption{Venn diagram of all states and a trace.}
\label{fig:states}
\end{figure}

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the CXL
machine works.
In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
at any point of time the CXL machine is in a specific state.
A state is comprised of the values of the shared variables as well as
the values of the process variables for each process, including its
program counter and the contents of its stack.
Everytime a process executes a CXL machine instruction, the
state changes (if only because the program counter of the process
changes).  We call that a \emph{step}.  Steps in CXL are atomic.

The CXL machine starts in an initial state in which there is only
one process and its program counter is~0.  A \emph{trace} is a
sequence of steps starting from the initial state.
When making a step, there are two sources of non-determinism in CXL.
One is when
there is more than one process that can make a step.  The other is
when a process executes a \texttt{choose} operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable} if it is either the initial state
or it can be reached from the initial state through a trace.
A state is final
when there are no processes left to make state changes.

A \emph{property} describes a subset of
states.  In other words, a property is equivalent to a subset of the states.
\emph{Initial}, \emph{final}, and \emph{reachable}, and \emph{unreachable}
are all examples of properties.
Figure~\ref{fig:states} depicts a Venn diagram of various properties
and a trace.
An \emph{invariant} is a property that holds over all reachable states.
We want to show that mutual exclusion is an invariant.
In other words, we want to show that the set of reachable states is a subset
of the set of states where there is at most one process in the critical
section.

One way to prove that a property is an invariant is through induction
on the number of steps.  First you prove that the property holds over
the initial state.  Then you prove that for every reachable state,
and for every step from that reachable state, the property also holds
over the resulting state.
For this you need a predicate that describes exactly which
states are reachable.
But we do not have such a predicate: we know how to describe the set
of reachable states, but given an arbitrary state it is not easy to
see whether it is reachable or not.

To solve this problem, we will use what is called an
\emph{inductive invariant}.
An inductive invariant $I$ is a predicate over states that satisfies the following:
\begin{itemize}
\item $I$ holds in the initial state.
\item For any state in which $I$ holds and any process in the
state takes a step, then $I$ also holds in the resulting state.
\end{itemize}
Unlike an invariant, an inductive invariant must also hold over unreachable states.

@@

So we are looking for an inductive invariant that \emph{implies} mutual
exclusion.  In other words, the set of states where the inducive invariant holds
must be a subset of the set of states where there is at most one process in
the critical section.
Let $C(i) = \mathtt{flags}[1 - i] \land
\mathtt{turn} = 1 - i$, that is, the condition on the \texttt{while} loop
for process~$i$.
Let predicate $I_p(i)$ be the following.
if process $i$ is label \texttt{@cs} (i.e., process $i$ is in the critical section),
then $C(i)$ does not hold or process $1-i$ is executing after setting
$\mathtt{flags}[i]$ but still before setting \texttt{turn} to $1-i$.
More formally, $I_p(i) = \mathtt{process}(i)@cs) \Rightarrow (\lnot C(i) \lor \mathtt{process}(1-i)@gate)$.
For Peterson's Algorithm, an inductive invariant that works well is
$I_p(0) \land I_p(1)$.
